Evolution of pre-copulatory and post-copulatory inbreeding behaviour and associated

### 2017-02-24 ms_pre_post.Rmd
Edits up through the last section of the Results

<!--- Jane, we might want to step back a bit on this one and figure out what exactly it is that is confusing and needs clarifying, as I'm not sure I see what the issue is anymore or, rather, why we're going into so much detail about a very routine procedure in an IBM (or any bit of code, really). This kind of sampling procedure was used in both of our previous IBM papers and I don't recall it being an issue. I have explained the procedure as carefully as I can, but it's getting to the point now where it's essentially turning the specifics of the C programming language into English (i.e., the methods here are now language dependent -- could be done differently in R or Perl or whatever, essentially just to say that we're sampling an element randomly from a weighted vector (I'm still not sure why we can't go back to just saying it that way; actually, I don't see why we can't just stop after ''the value of q^m_ij then defines the probability that a focal female mates with the male'' -- why does this need further explanation in the text?)). --->

@@

The probability that an offspring of a focal female $i$ is sired by any one of $i$'s mates $j$ is calculated by assigning a perceived fertilisation quality to each $j$ with respect to $i$, $Q^{f}_{ij}$. Perceived fertilisation quality $Q^{f}_{ij}$ is calculated in the same way as perceived mate quality $Q^{m}_{ij}$, such that if a focal female has a strategy of post-copulatory inbreeding avoidance ($F_{p}<0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = (-F_{p} \times k_{ij} + 1)^{-1}$. If the focal female has strategy of post-copulatory inbreeding preference ($F_{p}>0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = F_{p} \times k_{ij} + 1$. A relative quality ($q^{f}_{ij}$) is then calculated for each male by dividing his $Q^{f}_{ij}$ by the sum of the $Q^{f}_{ij}$ values across all of a female's mates. These $q^{f}_{ij}$ values, which lie between zero and one, define the probability of paternity. As with mate choice, $q^{f}_{ij}$ values are placed into a vector whose elements sum to unity, and a random uniform number is used to sample paternity. Females produce $n$ offspring, so $q^{f}_{ij}$ values are sampled $n$ times independently and with replacement for each female to determine the realised distribution of sires.

@@

@@

To address how asymmetric costs associated with pre-copulatory and post-copulatory inbreeding avoidance and polyandry affect the long-term persistence of reproductive strategies, we quantify change in $M_{a}$ and $F_{a}$ over generations in simulations where the cost of pre-copulatory inbreeding strategy $c_{M}=0$ but the cost of post-copulatory inbreeding strategy $c_{F}=0.02$, and vice versa. We compare these simulations with evolution of a costly strategy in the absence of evolution of an alternative strategy (e.g., evolution of pre-copulatory inbreeding strategy when post-copulatory inbreeding strategy phenotype is fixed at zero, $F_{p}=0$). In Supporting Information, we present results for all possible cost combinations of $0$ and $0.02$.

To address how selection on an initially rare strategy of pre-copulatory or post-copulatory inbreeding avoidance is affected by the existence of the other strategy of inbreeding avoidance already being fixed, we first used exploratory simulations to quantify evolution of pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy and associated polyandry, in isolation. Then, to test whether pre-copulatory inbreeding avoidance would evolve when adaptive polyandry and post-copulatory inbreeding avoidance was already established, we initiated $M_{a}$ allele values at zero, but fixed $F_{a}$ allele values at $-10$ and $P_{a}$ allele values at $1$ (i.e., meaning that $F_{p}$ and $P_{p}$ did not evolve). Similarly, to test whether post-copulatory inbreeding avoidance would evolve when pre-copulatory inbreeding avoidance was already established, we initiated $F_{a}$ and $P_{a}$ allele values at zero but fixed $M_{a}$ allele values at $-10$. Consequently, because females have 10 diploid loci, when $M_{a}$ or $F_{a}$ alleles were fixed at $-10$, outbred females were $51$ times less likely to choose a full brother and $13.5$ times times less likely to choose a first cousin than a non-relative in pre-copulatory and post-copulatory choice, respectively.

@@

When post-copulatory inbreeding avoidance alleles ($F_{a}$) were fixed to zero, meaning that $F_{p}$ could not evolve, $P_{a}$ alleles underlying polyandry always decreased to negative values over generations (Figure 1A,C,E,G). This shows that despite strong inbreeding depression in offspring viability, there is selection against unconditional polyandry even given zero direct cost ($c_{P}=0$; Figure 1A). This is because values of the $M_{a}$ also evolved to be negative, meaning that females typically avoided inbreeding through their initial mating. Polyandrous females that subsequently sampled more males from the available population were consequently more likely to mate with some relatives and hence produce some inbred offspring with low viability [@Duthie].

When post-copulatory inbreeding avoidance was allowed to evolve, mean $P_{a}$ values became substantially higher than in comparable simulations where $F_{a}$ values were fixed to zero and post-copulatory inbreeding avoidance could not evolve (Figure 1B,D,F,H). Allowing evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance therefore facilitated evolution of polyandry. However, $P_{a}$ values were expected to be positive only given a low cost of polyandry ($c_{P} < 0.005$, Figure 1B,D). This implies that, even given strong inbreeding depression in offspring viability, inbreeding avoidance might generally be a weak force driving polyandry evolution.

Strong post-copulatory inbreeding avoidance, manifested as very negative $F_{a}$ values, consistently evolved in all simulations where such evolution was allowed (Figure 1B,D,F,H). Evolution of post-copulatory inbreeding avoidance occurred even when $P_{a}$ values were expected to be slightly negative, and hence when there was selection against polyandry, (Figure 1F,H). This reflects the threshold nature of phenotypic expression of polygenic polyandry, wherein random sampling of alleles means that polyandry is commonly expressed (i.e., $P_{g}>0$) even when mean $P_{a}$ values are negative (Figure 2). This means that, even in populations where female reproductive strategy evolves toward monandry, there is still commonly some opportunity for expression of post-copulatory inbreeding avoidance and associated selection that drives evolution of $F_{a}$ allele values.

@@

When either pre-copulatory ($M_{a}$) or post-copulatory ($F_{a}$) inbreeding avoidance allele values were fixed to zero, the alternative inbreeding strategy evolved to inbreeding avoidance even when costly (Figure 3A,C). Females therefore evolved to avoid inbreeding, and thereby avoid the cost of producing inbred offspring, through whichever route was available.

However, when both pre-copulatory and post-copulatory inbreeding avoidance could evolve, their relative evolutionary dynamics depended on their relative costs. When one inbreeding strategy had zero cost, the other strategy initially evolved towards inbreeding avoidance (i.e., negative $M_{a}$ or $F_{a}$) but then evolved back towards random mating or fertilisation (i.e., $M_{a}=0$ or $F_{a}=0$) across generations following increasing evolution of inbreeding avoidance through the cost-free strategy (Figure 3B,D).

Evolution of post-copulatory inbreeding avoidance facilitated evolution of polyandry (Figure 3B,C), while monandry evolved in the absence of post-copulatory inbreeding avoidance (Figure 3A,D; polyandry was cost free in all these simulations).

@@

![Mean allele values underlying tendency for polyandry (red), pre-copulatory inbreeding strategy (blue), and post-copulatory inbreeding strategy (black) when (A and B) costly pre-copulatory inbreeding strategy can evolve and post-copulatory inbreeding strategy is (A) fixed for random fertilisation or (B) can also evolve, and when (C and D) costly post-copulatory inbreeding strategy can evolve and pre-copulatory inbreeding strategy is (A) fixed for random mating or (B) can also evolve. Mean allele values (solid lines) and associated standard errors (shading) are calculated across all individuals within a population over 40000 generations across 40 replicate populations. Negative mean allele values reflect strategies of inbreeding avoidance or tendency for monandry, and positive values reflect strategies of inbreeding preference or tendency for polyandry.](figures/rel_costs.png)

>>>>

<!--- We might want to step back a bit on this one and figure out what exactly it is that is confusing and needs clarifying, as I'm not sure I see what the issue is anymore or, rather, why we're going into so much detail about a very routine procedure in an IBM (or any bit of code, really). This kind of sampling procedure was used in both of our previous IBM papers and I don't recall it being an issue. I have explained the procedure as carefully as I can, but it's getting to the point now where it's essentially turning the specifics of the C programming language into English (i.e., the methods here are now language dependent -- could be done differently in R or Perl or whatever, essentially just to say that we're sampling an element randomly from a weighted vector (I'm still not sure why we can't go back to just saying it that way; actually, I don't see why we can't just stop after ''the value of q^m_ij then defines the probability that a focal female mates with the male'' -- why does this need further explanation in the text?)). --->

@@

The probability that an offspring of a focal female $i$ is sired by any one of $i$'s mates $j$ is calculated by assigning a perceived fertilisation quality to each $j$ with respect to $i$, $Q^{f}_{ij}$. Perceived fertilisation quality $Q^{f}_{ij}$ is calculated in the same way as perceived mate quality $Q^{m}_{ij}$, such that if a focal female has a strategy of post-copulatory inbreeding avoidance ($F_{p}<0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = (-F_{p} \times k_{ij} + 1)^{-1}$. If the focal female has strategy of post-copulatory inbreeding preference ($F_{p}>0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = F_{p} \times k_{ij} + 1$. A relative quality ($q^{f}_{ij}$) is then calculated for each male by dividing his $Q^{f}_{ij}$ by the sum of the $Q^{f}_{ij}$ values across all of a female's mates. These $q^{f}_{ij}$ values, which lie between zero and one, define the probability of paternity. As with mate choice, one vector element representing a single male is samplied as weighted by these probabilities to determine paternity. Females produce $n$ offspring, so $q^{f}_{ij}$ values are sampled $n$ times independently and with replacement for each female to determine the realised distribution of sires.

<!--- I've tried another way of explaining that sampling is done from a vector of probabilities in the second to last sentence here --->

@@

<!--- See the PDF for the equation. Not sure why it's not showing up in the DOCX. --->

@@

To address how asymmetric costs associated with pre-copulatory and post-copulatory inbreeding avoidance and polyandry affect the long-term persistence of reproductive strategies, we quantify the change in $M_{a}$ and $F_{a}$ over generations in simulations where the cost of pre-copulatory inbreeding strategy $c_{M}=0$ but the cost of post-copulatory inbreeding strategy $c_{F}=0.02$, and vice versa. We compare these simulations with evolution of a costly strategy in the absence of evolution of an alternative strategy (e.g., evolution of pre-copulatory inbreeding strategy when post-copulatory inbreeding strategy phenotype is fixed at zero, $F_{p}=0$). In Supporting Information, we present results for all possible cost combinations of $0$ and $0.02$. Previous modelling using similar genetic architecture suggests that a cost of $0.02$ imposes strong but not overhwleming direct negative selection on polyandry [@Duthie].

To address how selection on an initially rare strategy of pre-copulatory or post-copulatory inbreeding avoidance is affected by the other strategy of inbreeding avoidance already being fixed, we first used exploratory simulations to quantify evolution of pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy and associated polyandry, in isolation. Then, to test whether pre-copulatory inbreeding avoidance would evolve when adaptive polyandry and post-copulatory inbreeding avoidance were fixed, we initiated $M_{a}$ allele values at zero, but fixed $F_{a}$ allele values at $-10$ and $P_{a}$ allele values at $1$ (i.e., meaning that $F_{p}$ and $P_{p}$ did not evolve). Similarly, to test whether post-copulatory inbreeding avoidance would evolve when pre-copulatory inbreeding avoidance was already fixed, we initiated $F_{a}$ and $P_{a}$ allele values at zero but fixed $M_{a}$ allele values at $-10$. Consequently, because females have 10 diploid loci, when $M_{a}$ or $F_{a}$ alleles were fixed at $-10$, outbred females were $51$ times less likely to choose a full brother and $13.5$ times times less likely to choose a first cousin than a non-relative in pre-copulatory and post-copulatory choice, respectively.

@@

When post-copulatory inbreeding avoidance alleles ($F_{a}$) were fixed to zero, meaning that $F_{p}$ could not evolve, $P_{a}$ alleles underlying polyandry always decreased to negative values over generations (red lines, Figure 1A,C,E,G). This shows that despite strong inbreeding depression in offspring viability, there is selection against unconditional polyandry even given zero direct cost ($c_{P}=0$; Figure 1A). This is because $M_{a}$ values became negative over generations, meaning that females typically avoided inbreeding through their initial mating (blue lines, Figure 1A,C,E,G). Polyandrous females that subsequently sampled more males from the available population were consequently more likely to mate with some relatives and hence produce some inbred offspring with low viability [@Duthie].

When post-copulatory inbreeding avoidance was allowed to evolve, mean $P_{a}$ values became substantially higher than in comparable simulations where $F_{a}$ values were fixed to zero and post-copulatory inbreeding avoidance could not evolve (Figure 1B,D,F,H). Allowing evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance therefore facilitated evolution of polyandry. However, $P_{a}$ values were expected to become positive only given a low cost of polyandry ($c_{P} < 0.005$, Figure 1B,D versus 1F,H). This implies that, even given strong inbreeding depression in offspring viability, inbreeding avoidance might generally be a weak force driving polyandry evolution.

<!--- I've not added the suggested ''to the degree that most females mated multiply'' because it's not actually true unless costs are less than or equal to 0.025. Also, do we really need to point out figure 2 here? The very next paragraph explains that the degree of polyandry isn't trivial, and I should hope that readers aren't so impatient or put off by Figure 1 that they stop reading entirely!  --->

Strong post-copulatory inbreeding avoidance, manifested as very negative $F_{a}$ values, consistently evolved in all simulations where such evolution was allowed (black lines, Figure 1B,D,F,H). Evolution of post-copulatory inbreeding avoidance occurred even when $P_{a}$ values were expected to be slightly negative, and hence when there was selection against polyandry, (Figure 1F,H). This reflects the threshold nature of phenotypic expression of polygenic polyandry, wherein random sampling of alleles means that polyandry is expressed expressed by some females (i.e., $P_{g}>0$) even when mean $P_{a}$ values are negative (Figure 2C,D). This means that, even in populations where female reproductive strategy evolves toward monandry, there is still commonly some opportunity for expression of post-copulatory inbreeding avoidance and associated selection that drives evolution of post-copulatory inbreeding avoidance.

<!--- I'm not sure what aspects of Figure 1 you want me to mention in the main text? Are these things that are also in the legend, and if so, do we really need to repeat them? --->

@@

When post-copulatory inbreeding avoidance allele values ($F_{a}$) were fixed to zero, pre-copulatory inbreeding avoidance evolved even when costly (Figure 3A). Likewise, when pre-copulatory inbreeding avoidance allele values ($M_{a}$) were fixed to zero, costly post-copulatory inbreeding avoidance evolved (Figure 3C). Females therefore evolved to avoid inbreeding, and thereby avoid the cost of producing inbred offspring, through whichever route was available.

<!--- Last time we were through this, it was suggested that I scrap the former term ''inbreeding strategy'' because we don't deal at all with inbreeding preference. I have rewritten these sections in attempt to make the writing clearer. TODO: Additionally, I have tweaked the figure legends appropriately.--->

However, when both pre-copulatory and post-copulatory inbreeding avoidance could evolve, their relative evolutionary dynamics depended on their relative costs. When pre-copulatory but not post-copulatory inbreeding avoidance was costly ($c_{M}=0.02$ and $c_{F}=0$), pre-copulatory inbreeding avoidance initially evolved (i.e., negative $M_{a}$) but then evolved back towards random mating (i.e., $M_{a}=0$) following increasing evolution of post-copulatory inbreeding avoidance (Figure 3B). Similarly, when post-copulatory but not pre-copulatory inbreeding avoidance was costly ($c_{F}=0.02$ and $c_{M}=0$), post-copulatory inbreeding avoidance intially evolved ($F_{a}=0$) before returning to random fertilisation after ca $20000$ generations (Figure 3D).

Evolution of post-copulatory inbreeding avoidance facilitated evolution of polyandry (Figure 3B,C). In the absence of post-copulatory inbreeding avoidance, polyandry did not evolve (Figure 3A,D; polyandry was cost free in all these simulations).

@@

![Mean allele values underlying tendency for polyandry (red), pre-copulatory inbreeding strategy (blue), and post-copulatory inbreeding strategy (black) when (A and B) costly pre-copulatory inbreeding strategy can evolve and post-copulatory inbreeding strategy is (A) fixed for random fertilisation or (B) can also evolve, and when (C and D) costly post-copulatory inbreeding strategy can evolve and pre-copulatory inbreeding strategy is (A) fixed for random mating or (B) can also evolve. Mean allele values (solid lines) and associated standard errors (shading) are calculated across all individuals within a population over 40000 generations across 40 replicate populations. Negative mean allele values reflect strategies of inbreeding avoidance or tendency for monandry, and positive values reflect strategies of inbreeding preference or tendency for polyandry. In all panels, polyandry is cost free.](figures/rel_costs.png)
